---
title: "datatype-benchmark-rdgidb"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{datatype-benchmark-rdgidb}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

This vignette compares the microbenchmarks of get_drugs including:

- The original implementation (Direct port from dgipy)
- New implementations
- A data table implementation
- A tibble implementation (Following tidyverse principles)
- A tibblify implementation

A data frame implementation was considered but ultimately rejected as tibbles are the modern extension of data frames that generally more popular. Since the two are largely interchangeable, tibbles are favored for their enhanced functionality and user-friendliness.

## Load Packages

```{r setup, message=FALSE}
library(microbenchmark)
library(httptest)

library(data.table)
library(tibble)
library(purrr)
library(dplyr)
library(tibblify)
```

```{r, include = FALSE}
httptest::start_vignette("datatype-benchmark-rdgidb")
```

## Helper Functions

```{r helper}
group_attributes <- function(row) {
  grouped_dict <- list()
  for (attr in row) {
    if (is.null(attr$value)) {
      next
    } else if (attr$name %in% names(grouped_dict)) {
      grouped_dict[[attr$name]] <- append(grouped_dict[[attr$name]], attr$value)
    } else {
      grouped_dict[[attr$name]] <- list(attr$value)
    }
  }
  grouped_dict
}
 
backfill_dicts <- function(col) {
  keys <- unique(unlist(lapply(col, names)))
  result <- lapply(col, function(cell) sapply(keys, function(key) cell[[key]]))
  result
}

get_drugs_query <- function(
    terms,
    immunotherapy = NULL,
    antineoplastic = NULL,
    api_url = NULL) {
  params <- list(names = terms)
  if (!is.null(immunotherapy)) {
    params$immunotherapy <- immunotherapy
  }
  if (!is.null(antineoplastic)) {
    params$antineoplastic <- antineoplastic
  }

  api_url <- if (!is.null(api_url)) api_url else api_endpoint_url
  query_file_path <- system.file(
    "queries/get_drugs.graphql",
    package = "rdgidb"
  )
  query <- readr::read_file(query_file_path)
  response <- httr::POST(
    api_url,
    body = list(query = query, variables = params),
    encode = "json"
  )
  result <- httr::content(response)$data
  result
}
```

## Initalize Inputs

```{r inputs}
drugs <- c("imatinib", "metronidazole", "sunitinib", "zalcitabine", "trastuzumab")

api_endpoint_url <- Sys.getenv(
  "DGIDB_API_URL",
  unset = "https://dgidb.org/api/graphql"
)

result <- httptest::with_mock_api({
  get_drugs_query(drugs)
})
```

# Microbenchmarks

## Original Implementation

```{r original}
get_drugs_original <- function(result) {
  output <- list(
    drug_name = list(),
    drug_concept_id = list(),
    drug_aliases = list(),
    drug_attributes = list(),
    drug_is_antineoplastic = list(),
    drug_is_immunotherapy = list(),
    drug_is_approved = list(),
    drug_approval_ratings = list(),
    drug_fda_applications = list()
  )

  for (match in result$drugs$nodes) {
    output$drug_name <- append(
      output$drug_name, match$name
    )
    output$drug_concept_id <- append(
      output$drug_concept_id, match$conceptId
    )
    output$drug_aliases <- append(
      output$drug_aliases, list(lapply(match$drugAliases, function(a) a$alias))
    )
    output$drug_attributes <- append(
      output$drug_attributes, list(group_attributes(match$drugAttributes))
    )
    output$drug_is_antineoplastic <- append(
      output$drug_is_antineoplastic, match$antiNeoplastic
    )
    output$drug_is_immunotherapy <- append(
      output$drug_is_immunotherapy, match$immunotherapy
    )
    output$drug_is_approved <- append(
      output$drug_is_approved, match$approved
    )
    output$drug_approval_ratings <- append(
      output$drug_approval_ratings,
      list(lapply(match$drugApprovalRatings, function(r) {
        list(rating = r$rating, source = r$source$sourceDbName)
      }))
    )
    output$drug_fda_applications <- append(
      output$drug_fda_applications,
      list(lapply(match$drugApplications, function(app) app$appNo))
    )
  }
  output$drug_attributes <- backfill_dicts(output$drug_attributes)
  output
}
```

```{r original-benchmark}
microbenchmark(
  get_drugs_original(result),
  times = 1000
)
```

## New (Speed) Implementation

```{r new-speed}
get_drugs_new_speed <- function(result) {
  nodes <- result$drugs$nodes
  n <- length(nodes)

  drug_name <- character(n)
  drug_concept_id <- character(n)
  drug_aliases <- vector("list", n)
  drug_attributes <- vector("list", n)
  drug_is_antineoplastic <- logical(n)
  drug_is_immunotherapy <- logical(n)
  drug_is_approved <- logical(n)
  drug_approval_ratings <- vector("list", n)
  drug_fda_applications <- vector("list", n)

  for (i in seq_len(n)) {
    node <- nodes[[i]]
    drug_name[i] <- node$name
    drug_concept_id[i] <- node$conceptId
    drug_aliases[[i]] <- vapply(node$drugAliases, function(a) a$alias, character(1))
    drug_attributes[[i]] <- group_attributes(node$drugAttributes)
    drug_is_antineoplastic[i] <- node$antiNeoplastic
    drug_is_immunotherapy[i] <- node$immunotherapy
    drug_is_approved[i] <- node$approved
    drug_approval_ratings[[i]] <- lapply(node$drugApprovalRatings, function(r) {
      list(rating = r$rating, source = r$source$sourceDbName)
    })
    drug_fda_applications[[i]] <- vapply(node$drugApplications, function(a) a$appNo, character(1))
  }
  drug_attributes <- backfill_dicts(drug_attributes)

  output <- list(
    drug_name = drug_name,
    drug_concept_id = drug_concept_id,
    drug_aliases = drug_aliases,
    drug_attributes = drug_attributes,
    drug_is_antineoplastic = drug_is_antineoplastic,
    drug_is_immunotherapy = drug_is_immunotherapy,
    drug_is_approved = drug_is_approved,
    drug_approval_ratings = drug_approval_ratings,
    drug_fda_applications = drug_fda_applications
  )
  output
}
```

```{r new-speed-benchmark}
microbenchmark(
  get_drugs_new_speed(result),
  times = 1000
)
```

## New (Compact) Implementation

```{r new-compact}
get_drugs_new_compact <- function(result) {
  nodes <- result$drugs$nodes
  output <- list(
    drug_name = vapply(nodes, function(x) x$name, character(1)),
    drug_concept_id = vapply(nodes, function(x) x$conceptId, character(1)),
    drug_aliases = lapply(nodes, function(x) vapply(x$drugAliases, function(a) a$alias, character(1))),
    drug_attributes = lapply(nodes, function(x) group_attributes(x$drugAttributes)),
    drug_is_antineoplastic = vapply(nodes, function(x) x$antiNeoplastic, logical(1)),
    drug_is_immunotherapy = vapply(nodes, function(x) x$immunotherapy, logical(1)),
    drug_is_approved = vapply(nodes, function(x) x$approved, logical(1)),
    drug_approval_ratings = lapply(nodes, function(x) {
      lapply(x$drugApprovalRatings, function(r) list(rating = r$rating, source = r$source$sourceDbName))
    }),
    drug_fda_applications = lapply(nodes, function(x) vapply(x$drugApplications, function(a) a$appNo, character(1)))
  )
  output$drug_attributes <- backfill_dicts(output$drug_attributes)
  output
}
```

```{r new-compact-benchmark}
microbenchmark(
  get_drugs_new_compact(result),
  times = 1000
)
```

## As Tibble Implementation

```{r as-tibble}
get_drugs_as_tibble <- function(result) {
  output <- get_drugs_new_speed(result)
  output <- tibble::as_tibble(output)
  output
}
```

```{r as-tibble-benchmark}
microbenchmark(
  get_drugs_as_tibble(result),
  times = 1000
)
```

## Data Table Implementation

```{r datatable}
get_drugs_datatable <- function(result) {
  nodes <- result$drugs$nodes
  node_list <- lapply(nodes, function(x) list(
    drug_name = x$name,
    drug_concept_id = x$conceptId,
    drug_aliases = list(vapply(x$drugAliases, function(a) a$alias, character(1))),
    drug_attributes = list(group_attributes(x$drugAttributes)),
    drug_is_antineoplastic = x$antiNeoplastic,
    drug_is_immunotherapy = x$immunotherapy,
    drug_is_approved = x$approved,
    drug_approval_ratings = list(lapply(x$drugApprovalRatings, function(r) list(
      rating = r$rating,
      source = r$source$sourceDbName
    ))),
    drug_fda_applications = list(vapply(x$drugApplications, function(a) a$appNo, character(1)))
  ))
  output <- data.table::rbindlist(node_list, fill = TRUE)
  output[, drug_attributes := list(backfill_dicts(drug_attributes))]
  output
}
```

```{r datatable-benchmark}
microbenchmark(
  get_drugs_datatable(result),
  times = 1000
)
```

## Tibble Implementation

```{r tibble}
get_drugs_tibble <- function(result) {
  nodes <- result$drugs$nodes
  tibble::tibble(
    drug_name = purrr::map_chr(nodes, "name"),
    drug_concept_id = purrr::map_chr(nodes, "conceptId"),
    drug_aliases = purrr::map(nodes, ~purrr::map_chr(.x$drugAliases, "alias")),
    drug_attributes = purrr::map(nodes, ~group_attributes(.x$drugAttributes)),
    drug_is_antineoplastic = purrr::map_lgl(nodes, "antiNeoplastic"),
    drug_is_immunotherapy = purrr::map_lgl(nodes, "immunotherapy"),
    drug_is_approved = purrr::map_lgl(nodes, "approved"),
    drug_approval_ratings = purrr::map(
      nodes,
      ~purrr::map(.x$drugApprovalRatings, ~list(rating = .x$rating, source = .x$source$sourceDbName))
    ),
    drug_fda_applications = purrr::map(nodes, ~purrr::map_chr(.x$drugApplications, "appNo"))
  ) %>%
  dplyr::mutate(
    drug_attributes = backfill_dicts(drug_attributes)
  )
}
```

```{r tibble-benchmark}
microbenchmark(
  get_drugs_tibble(result),
  times = 1000
)
```

## Tibblify Implementation

```{r tibblify}
get_drugs_tibblify <- function(result) {
  tspec <- tibblify::tspec_df(
    drug_name = tibblify::tib_chr("name"),
    drug_concept_id = tibblify::tib_chr("conceptId"),
    drug_aliases  = tibblify::tib_df(
      "drugAliases",
      rating = tibblify::tib_chr("alias"),
    ),
    drug_attributes = tibblify::tib_df(
      "drugAttributes",
      attribute_name  = tibblify::tib_chr("name"),
      attribute_value = tibblify::tib_chr("value")
    ),
    drug_is_antineoplastic = tibblify::tib_lgl("antiNeoplastic"),
    drug_is_immunotherapy = tibblify::tib_lgl("immunotherapy"),
    drug_is_approved = tibblify::tib_lgl("approved"),
    drug_approval_ratings  = tibblify::tib_df(
      "drugApprovalRatings",
      rating = tibblify::tib_chr("rating"),
      source = tibblify::tib_chr("source.sourceDbName", required = FALSE)
    ),
    drug_fda_applications  = tibblify::tib_df(
      "drugApplications",
      rating = tibblify::tib_chr("appNo"),
    ),
  )
  output <- tibblify::tibblify(result$drugs$nodes, tspec)
  output
}
```

```{r tibblify-benchmark}
microbenchmark(
  get_drugs_tibblify(result),
  times = 1000
)
```