---
title: "rDGIdb: R Interface to the 'DGIdb' GraphQL API"
author:
  - name: Rohit Basu
date: "`r BiocStyle::doc_date()`"
package: "`r BiocStyle::pkg_ver('rDGIdb')`"
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{rDGIdb: R Interface to the 'DGIdb' GraphQL API}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    warning = FALSE,
    message = FALSE
)

online <- interactive()
```

# Introduction

The Drug-Gene Interaction Database (DGIdb) provides information about drugs,
genes, drug-gene interactions, and other therapeutically relevant information.
`rDGIdb` provides an R interface to the public DGIdb GraphQL API so that these
data can be queried directly from R and incorporated into Bioconductor
workflows.

The package provides a set of query functions that can pull information from
DGIdb.
Results are returned as named lists in a column-oriented format: each element
represents a field, and values at the same position describe the same record.

Project source and issue tracking are available at
<https://github.com/dgidb/r-dgidb>. The DGIdb web application is available at
<https://dgidb.org/>.

# Getting started

Load `rDGIdb` with:

```{r load-package}
library(rDGIdb)
```

The main functions are:

| Function | Purpose |
| --- | --- |
| `getGenes()` | Look up gene records |
| `getDrugs()` | Look up drug records |
| `getInteractions()` | Find drug-gene interactions by gene or drug |
| `getCategories()` | Retrieve DGIdb gene-category annotations |
| `getSources()` | Inspect DGIdb aggregate data sources |
| `getDrugApplications()` | Retrieve Drugs\@FDA product information |
| `getAllGenes()` | Retrieve all DGIdb gene names and concept identifiers |
| `getAllDrugs()` | Retrieve all DGIdb drug names and concept identifiers |

# Looking up genes

Use `getGenes()` to retrieve DGIdb records for one or more genes. For example,
the following query retrieves records for `BRAF` and `PDGFRA`:

```{r gene-lookup, eval = online}
genes <- getGenes(c("BRAF", "PDGFRA"))

gene_table <- data.frame(
    gene_name = genes$gene_name,
    gene_concept_id = genes$gene_concept_id
)

gene_table
```

Some fields contain multiple values per record. Gene aliases, for example, are
returned as a list column:

```{r gene-aliases, eval = online}
head(genes$gene_aliases[[1]])
```

# Looking up drug-gene interactions

A common use of `rDGIdb` is to start with genes of interest and retrieve known
or reported drug-gene interactions. By default, `getInteractions()` interprets
its input as gene names:

```{r gene-interactions, eval = online}
interactions <- getInteractions(c("BRAF", "PDGFRA"))

interaction_table <- data.frame(
    gene_name = interactions$gene_name,
    drug_name = interactions$drug_name,
    drug_approved = interactions$drug_approved,
    interaction_score = interactions$interaction_score
)

head(interaction_table)
```

Interactions can also be queried starting from drug names by setting
`search = "drugs"`:

```{r drug-interactions, eval = online}
imatinib_interactions <- getInteractions(
    "Imatinib",
    search = "drugs",
    approved = TRUE
)

head(data.frame(
    gene_name = imatinib_interactions$gene_name,
    drug_name = imatinib_interactions$drug_name,
    interaction_score = imatinib_interactions$interaction_score
))
```

Additional filters are available for source database, PubMed identifier,
interaction type, approval status, immunotherapy status, and antineoplastic
use. Refer to `?getInteractions` for the complete set of arguments.

# Looking up drugs

Use `getDrugs()` to retrieve drug records. The returned fields include DGIdb
concept identifiers, aliases, attributes, approval information, and associated
Drugs\@FDA application identifiers.

```{r drug-lookup, eval = online}
drugs <- getDrugs("Imatinib")

data.frame(
    drug_name = drugs$drug_name,
    drug_concept_id = drugs$drug_concept_id,
    approved = drugs$drug_is_approved,
    antineoplastic = drugs$drug_is_antineoplastic,
    immunotherapy = drugs$drug_is_immunotherapy
)
```

Drug searches can also be restricted by antineoplastic or immunotherapy status:

```{r drug-filter, eval = online}
getDrugs("Imatinib", antineoplastic = TRUE)
```

# Gene categories

DGIdb annotates genes with categories that can help identify genes of potential
therapeutic interest. Use `getCategories()` to retrieve these annotations:

```{r categories, eval = online}
categories <- getCategories("BRAF")

data.frame(
    gene_name = categories$gene_name,
    gene_category = categories$gene_category
)
```

# DGIdb data sources

DGIdb aggregates information from multiple source databases. `getSources()`
returns information about these sources, including their names, versions,
claim counts, and licensing information.

The `sourceTypes` object lists the source categories accepted by
`getSources()`:

```{r source-types}
sourceTypes
```

For example, interaction sources can be queried with:

```{r sources, eval = online}
sources <- getSources(sourceTypes$INTERACTION)

head(data.frame(
    source_name = sources$source_name,
    source_version = sources$source_version,
    interaction_claims = sources$source_interaction_claims
))
```

# Drugs\@FDA application information

`getDrugApplications()` combines DGIdb drug application identifiers with
product information from the openFDA Drugs\@FDA API.

```{r fda-applications, eval = online}
applications <- getDrugApplications("Imatinib")

head(data.frame(
    drug_name = applications$drug_name,
    application = applications$drug_product_application,
    brand_name = applications$drug_brand_name,
    dosage_form = applications$drug_dosage_form,
    dosage_strength = applications$drug_dosage_strength
))
```

# Retrieving complete gene or drug lists

`getAllGenes()` and `getAllDrugs()` retrieve all gene or drug names and concept
identifiers currently available through DGIdb.

```{r all-records, eval = FALSE}
all_genes <- getAllGenes()
all_drugs <- getAllDrugs()
```

# Session information

```{r session-info}
sessionInfo()
```
